# (PART) DATA PROCESSING {-}

# Processing `mothur` output in R

## Import library and functions
```{r}
source("R/common.R")
source("R/common.R")
library(tidyverse, suppressPackageStartupMessages())
set.seed(110912)
```


## Creating `mothur` otu and taxonomy tables
```{r}
library(tidyverse)
mo_otutable_long <- read_tsv("data/final.opti_mcc.shared", show_col_types = FALSE) %>% 
  select(-label, -numOtus) %>% 
  rename(sample_id = Group) %>% 
  rename_all(tolower) %>% 
  pivot_longer(-sample_id, names_to="otu", values_to="count") %>% 
  filter(count != 0)

mo_otutable <- mo_otutable_long %>%
  pivot_wider(id_cols = otu, names_from = sample_id, values_from = count) %>%
  mutate_if(is.numeric, ~replace(., is.na(.), 0))

cat("Dimension of the otutable\n")
dim(mo_otutable)
cat("\nStructure of the otutable\n")
head(mo_otutable)[, 1:3] %>% as.data.frame()  
saveRDS(mo_otutable, "RDataRDS/mo_otutable.rds")

taxlevels <- c("kingdom", "phylum", "class", "order", "family", "genus")
mo_taxonomy <- read_tsv("data/final.opti_mcc.0.03.cons.taxonomy", show_col_types = F) %>%
  rename_all(tolower) %>%
  select(otu, taxonomy) %>%
  mutate(otu = tolower(otu),
         taxonomy = str_replace_all(taxonomy, "\\(\\d+\\)", ""),
         taxonomy = str_replace(taxonomy, ";unclassified", "_unclassified"),
         taxonomy = str_replace_all(taxonomy, ";unclassified", ""),
         taxonomy = str_replace_all(taxonomy, ";$", "")) %>% 
  separate(taxonomy, into = all_of(taxlevels), sep = ";")

cat("Dimension of the taxonomy table\n")
dim(mo_taxonomy)
cat("\nStructure of the taxonomy table\n")
head(mo_taxonomy)[, 1:4] %>% as.data.frame() 
saveRDS(mo_taxonomy, "RDataRDS/mo_taxonomy.rds")

save(mo_otutable, mo_taxonomy, file = "RDataRDS/otutaxonomy_objects.RData")
resave(mo_otutable, mo_taxonomy, file = "~/Dropbox/CDILLC/GIT_REPOS/smda-end2end/RDataRDS/saved_objects.RData")
```

# Processing `qiime2` Output in R

## Creating `qiime2` taxa abundance table {#mo-otu-taxonomy}
- Here we read through the merged feature-taxonomy-abundance table and extract abundance at a taxon lineage.
- The demo below extract and split abundance at genus-lineage.
- Additional processing is also done to create a tidy dataframe.
- The final out put will contain feature ID, taxonomy and abundance for each sample.
```{r}
taxlevels <- c("kingdom", "phylum", "class", "order", "family", "genus")

q2_taxa_abund <- read_tsv("data/q2-transformed-tables/q2_feature_tax_abund.tsv", skip = 0, show_col_types = FALSE)[-1,] %>%
  relocate(Taxon, .after = id) %>% 
  mutate_at(3:ncol(.), as.numeric) %>% 
  rename(feature = id,
         taxonomy = Taxon) %>% 
  relocate(taxonomy, .after = feature) %>% 
  distinct() %>%
  mutate_if(is.numeric, ~replace(., is.na(.), 0)) %>%
  mutate(taxonomy = str_replace_all(taxonomy, "; s__$", ""),
         taxonomy = str_replace_all(taxonomy, "; g__$", ""),
         taxonomy = str_replace_all(taxonomy, "; f__$", ""),
         taxonomy = str_replace_all(taxonomy, "; o__$", ""),
         taxonomy = str_replace_all(taxonomy, "; c__$", ""),
         taxonomy = str_replace_all(taxonomy, "; p__$", ""),
         taxonomy = str_replace_all(taxonomy, "; k__$", ""),
         taxonomy = str_replace_all(taxonomy, "\\[|\\]", ""),
         taxonomy = str_replace_all(taxonomy, "\\s", "")) %>%
  dplyr::filter(!grepl("s__*", taxonomy)) %>%
  dplyr::filter(grepl("g__*", taxonomy)) %>% 
  select(-Confidence) %>% 
  mutate(taxonomy = str_replace_all(taxonomy, "\\w__", "")) %>% 
  separate(taxonomy, into = all_of(taxlevels), sep = ";")

colnames(q2_taxa_abund)
saveRDS(q2_taxa_abund, "RDataRDS/q2_taxa_abund.rds")


taxlevels <- c("kingdom", "phylum", "class", "order", "family", "genus")

q2_otutable <- q2_taxa_abund %>%
  select(-all_of(taxlevels))

cat("Dimension of the otutable\n")
dim(q2_otutable)
cat("\nStructure of the otutable\n")
head(q2_otutable)[, 1:3] %>% as.data.frame()  
saveRDS(q2_otutable, "RDataRDS/q2_otutable.rds")

q2_taxonomy <- q2_taxa_abund %>%
  rename_all(tolower) %>%
  select(feature, all_of(taxlevels)) %>%
  distinct()

cat("Dimension of the taxonomy table\n")
dim(q2_taxonomy)
cat("\nStructure of the taxonomy table\n")
head(q2_taxonomy)[, 1:3] %>% as.data.frame() 
saveRDS(q2_taxonomy, "RDataRDS/q2_taxonomy.rds")

resave(q2_otutable, q2_taxonomy, file = "RDataRDS/otutaxonomy_objects.RData")
resave(q2_otutable, q2_taxonomy, file = "~/Dropbox/CDILLC/GIT_REPOS/smda-end2end/RDataRDS/saved_objects.RData")
```


# Processing `metaphlan` output in R

## Merging `metaphlan` abundance files
- First we create an object to specify the location of `metaphlan` output.
- Then we an empty data frame containing an empty column for holding the file names.
- Then we loop through each file to extract the output and create a merged abundance table.
- We will then create a tidy table containing sample ID, taxonomy, relative abundance and read statistics.

> Note: We will specify the path where the `metaphlan` output is located. Replace the path to match yours.

```{r withcounts}
path <- "/Volumes/RedSeagate/SYNTHETIC/microbes/data/*_profile.txt"
files <- Sys.glob(path)

merged_files <- data.frame(
  ModelName = character(), 
  Object = character(), 
  stringsAsFactors = FALSE) %>%
  mutate(Filename = "")

for (i in 1:length(files)){
  currentFile = read_delim(files[i], skip = 4, show_col_types = FALSE) %>% 
    mutate(Filename = files[i])
    merged_files = rbind(merged_files, currentFile)}

m3_merged_abund_table <- merged_files %>% 
  rename_all(tolower) %>%
  rename(clade_name = colnames(merged_files)[1],
         sample_id = filename,
         rel_abund = relative_abundance, 
         mapped_read = estimated_number_of_reads_from_the_clade) %>% 
  mutate(sample_id = str_replace_all(sample_id, ".*/", ""),
         sample_id = str_replace_all(sample_id, ".fast.*", "")) %>%
  select(sample_id, taxonomy = clade_name, rel_abund, mapped_read)
    
colnames(m3_merged_abund_table)

saveRDS(m3_merged_abund_table, "RDataRDS/m3_merged_abund_table.rds")
resave(m3_merged_abund_table, file = "~/Dropbox/CDILLC/GIT_REPOS/smda-end2end/RDataRDS/saved_objects.RData")
```


## Creating `metaphlan` relative abundance and taxonomy tables {#m3-otu-taxonomy}
- Here we read through the merged abundance table and extract abundance at a taxon lineage.
- The demo below extract and split abundance at genus-lineage and add code each row to a unique OTU ID.
- Additional processing is also done to create a tidy dataframe.
- The final out put will contain OTU ID, taxonomy and abundance for each sample.

```{r}
taxlevels <- c("kingdom", "phylum", "class", "order", "family", "genus")

m3_taxa_abund <- readRDS("RDataRDS/m3_merged_abund_table.rds") %>% 
  select(sample_id, taxonomy, rel_abund) %>% 
  distinct() %>% 
  dplyr::filter(!grepl("\\|s__*", taxonomy),
                grepl("g__*", taxonomy)) %>%
  group_by(sample_id, taxonomy, rel_abund) %>%
  summarise(count = sum(rel_abund),
            rel_abund = rel_abund/100, .groups = "drop") %>% 
  pivot_wider(id_cols = "taxonomy", names_from = "sample_id", values_from = "rel_abund") %>% 
  mutate(otu = paste("otu", seq(1:length(rownames(.))), sep = ""), .before = 1) %>% 
  mutate_if(is.numeric, ~replace(., is.na(.), 0 )) %>% 
  mutate(taxonomy = str_replace_all(taxonomy, "\\w__", "")) %>%
  separate(taxonomy, into = all_of(taxlevels), sep = "\\|") 

m3_otutable <- m3_taxa_abund %>% 
  select(-all_of(taxlevels))

cat("Dimension of the otutable\n")
dim(m3_otutable)
cat("\nStructure of the otutable\n")
head(m3_otutable)[, 1:5] %>% as.data.frame()  
saveRDS(m3_otutable, "RDataRDS/m3_otutable.rds")

m3_taxonomy <- readRDS("RDataRDS/m3_taxa_abund.rds") %>% 
  select(otu, all_of(taxlevels))

cat("Dimension of the taxonomy table\n")
dim(m3_taxonomy)
cat("\nStructure of the taxonomy table\n")
head(m3_taxonomy)[, 1:4] %>% as.data.frame() 
saveRDS(m3_taxonomy, "RDataRDS/m3_taxonomy.rds")

resave(m3_otutable, m3_taxonomy, file = "RDataRDS/otutaxonomy_objects.RData")
resave(m3_otutable, m3_taxonomy, file = "~/Dropbox/CDILLC/GIT_REPOS/smda-end2end/RDataRDS/saved_objects.RData")
```


## Creating `metaphlan` raw and relative abundance based on read count {#m3-read-count_abund}
- Here we read through the merged abundance table and extract abundance at a taxon lineage.
- The demo below extract and split abundance at genus-lineage and add code each row to a unique OTU ID.
- Additional processing is also done to create a tidy dataframe.
- The final out put will contain OTU ID, taxonomy and abundance for each sample.

```{r}
taxlevels <- c("kingdom", "phylum", "class", "order", "family", "genus")

m3_read_count_taxa_abund <- readRDS("RDataRDS/m3_merged_abund_table.rds") %>% 
  select(sample_id, taxonomy, mapped_read) %>% 
  distinct() %>% 
  dplyr::filter(!grepl("\\|s__*", taxonomy),
                grepl("g__*", taxonomy)) %>%
  pivot_wider(id_cols = "taxonomy", names_from = "sample_id", values_from = "mapped_read") %>% 
  mutate(otu = paste("otu", seq(1:length(rownames(.))), sep = ""), .before = 1) %>% 
  mutate_if(is.numeric, ~replace(., is.na(.), 0 )) %>% 
  mutate(taxonomy = str_replace_all(taxonomy, "\\w__", "")) %>%
  separate(taxonomy, into = all_of(taxlevels), sep = "\\|") 

m3_read_counts_otutable <- m3_read_count_taxa_abund %>% 
  select(-all_of(taxlevels))

cat("Dimension of the otutable\n")
dim(m3_read_counts_otutable)
cat("\nStructure of the otutable\n")
head(m3_read_counts_otutable)[, 1:5] %>% as.data.frame()  
saveRDS(m3_read_counts_otutable, "RDataRDS/m3_read_counts_otutable.rds")

m3_taxonomy <- readRDS("RDataRDS/m3_taxa_abund.rds") %>% 
  select(otu, all_of(taxlevels))

cat("Dimension of the taxonomy table\n")
dim(m3_taxonomy)
cat("\nStructure of the taxonomy table\n")
head(m3_taxonomy)[, 1:4] %>% as.data.frame() 
saveRDS(m3_taxonomy, "RDataRDS/m3_taxonomy.rds")

resave(m3_read_counts_otutable, m3_taxonomy, file = "RDataRDS/otutaxonomy_objects.RData")
resave(m3_read_counts_otutable, m3_taxonomy, file = "~/Dropbox/CDILLC/GIT_REPOS/smda-end2end/RDataRDS/saved_objects.RData")
```


